---
title: "Lectures 6 & 7 Demo"
output: html_notebook
---

# Start to Finish Analysis for the Jackson Heart Study

For Homework 4, you will need to do an analyze a dataset to answer a research question. To prepare you for this homework, we will be running through an analysis of data form the [Jackson Heart Study](https://www.jacksonheartstudy.org/). You can use this analysis we do together as a template for your homework and for future analyses.

Our research question for this anaysis will be:

> Is smoking associated with hypertension after controlling for sociodemographic variables?

# Data Exploration

Library the packages you need and import `baseline_jhs.csv`. The dataset and the codebook are on Canvas and Ed.

```{r setup}
library(here)
library(tidyverse)

jhs_data = read_csv(here("data/baseline_jhs.csv"))
```

Explore the data structure and variables. Here are some questions you should be able to answer:

-   What does each row represent? - Patient

-   What does each column represent? - Variable

-   What is the shape of the continuous variables?

    -   Issue with age, observation(s) at 999

    -   Alc/week heavily right skewed

-   What does each group of the categorical variables represent?

    -   MaleV1 - Sex: Male or Female
    -   IncomeV1 - 4 categories of income: Poor, Lower-middle, Upper-middle, or Affluent
    -   edu3catV1 - 3 categories of education: \<HS, High School/GED, or Some college or more

-   Which variable(s) have a lot of missingness?

    -   Income is 14.5% missing

```{r}
# See part of the data
head(jhs_data)

# Continuous variable shape
hist(jhs_data$ageV1, breaks = 100)
summary(jhs_data$ageV1)
hist(jhs_data$alcwV1)

# Categorical variables
table(jhs_data$maleV1, useNA = "ifany")
table(jhs_data$IncomeV1, useNA = "ifany")
table(jhs_data$edu3catV1, useNA = "ifany")

# Missingness
prop.table(table(is.na(jhs_data$IncomeV1)))
mean(is.na(jhs_data$IncomeV1))
```

# Data Cleaning

Based on our findings from the data exploration above, let's made a plan for data cleaning. We ***always*** want to document how we are cleaning our data outside of the code so that even people who cannot read R code know what is happening to the data.

```{r}
# These two do the same thing
# 1) Preferred option. Assume anyone over 120 years old doesn't have a valid age
#     Use mutate and if_else to make all of their ages NA
clean_jhs_data = jhs_data %>%
  mutate(ageV1 = if_else(ageV1>120, NA, ageV1))

# 2) Creates two datasets. One is everyone with a valid age <120
#     The other is everyone over 120 years, that then makes all of their ages NA
#     We then use bind_rows to stack the two datasets
clean_jhs_data = jhs_data %>%
  filter(ageV1<120) %>%
  bind_rows(jhs_data %>%
              filter(ageV1>120) %>%
              mutate(ageV1=NA))
```

# Descriptive Statistics

## Categorical Variables

First, let's look at the relative counts of smoking and hypertension in our dataset.

```{r}
clean_jhs_data = clean_jhs_data %>%
  mutate(income2cat = case_when(IncomeV1=="Affluent" ~ "High",
                                IncomeV1=="Upper-middle" ~ "High",
                                is.na(IncomeV1) ~ NA,
                                T                        ~ "Low"))

table(clean_jhs_data$income2cat, clean_jhs_data$IncomeV1, useNA = "ifany")
```

What's our unadjusted odds for low income for smokers?

```{r}
table(clean_jhs_data$income2cat, clean_jhs_data$everSmokerV1)

258*1104/(447*426)
```

Let's look at the different distributions of socioeconomic variables like sex, income, and education across smoking.

```{r}
table(clean_jhs_data$maleV1, clean_jhs_data$everSmokerV1)

table(clean_jhs_data$edu3catV1, clean_jhs_data$everSmokerV1)

chisq.test(clean_jhs_data$maleV1, clean_jhs_data$everSmokerV1)
```

## Continuous Variables

Compare age across smoking and hypertension. Are the differences statistically significant?

```{r}
hist(clean_jhs_data$ageV1)

hist(clean_jhs_data %>% filter(maleV1=="Male") %>% pull(ageV1))
hist(clean_jhs_data %>% filter(maleV1=="Female") %>% pull(ageV1))

t.test(ageV1~maleV1, data=clean_jhs_data)
```

Compare depression score across smoking and hypertension. Are the differences statistically significant?

## Table 1

Let's make the Table 1 for our paper with our exposure (smoking) on the top and all of the other variables along the side. Try doing a few by hand before using `table1`.

```{r}
# Stats by hand
clean_jhs_data %>%
  group_by(everSmokerV1) %>%
  summarise(mean_age = mean(ageV1, na.rm=T),
            median_alc = median(alcwV1, na.rm=T),
            median_depression = median(depressionV1, na.rm=T))
table(clean_jhs_data$maleV1, clean_jhs_data$everSmokerV1, useNA = "ifany")
table(clean_jhs_data$IncomeV1, clean_jhs_data$everSmokerV1, useNA = "ifany")

# Using table1
library(table1)

# Filter out missing smoking variables b/c table1 won't accept them
table1_data = clean_jhs_data %>%
  filter(!is.na(everSmokerV1))

table1(~ageV1 + maleV1 + alcwV1 + IncomeV1 + edu3catV1 + dailyDiscrV1 + 
         depressionV1 + discrmBurdenV1 + sbpV1 + dbpV1 + income2cat | 
         everSmokerV1, data=table1_data)
```

------------------------------------------------------------------------

**Start Week 7**

# Statistical Modeling

## Unadjusted Model

Use a linear model to look at how smoking is associated with systolic blood pressure.

```{r}

```

What do the model parameters mean?

Use a logistic model to look at how smoking is associated with high blood pressure.

```{r}

```

What do the model parameters mean?

## Adjusted Model

Use a linear model to look at how smoking is associated with systolic blood pressure, controlling for sociodemographic variables.

```{r}

```

What do the model parameters mean?

Use a logistic model to look at how smoking is associated with high blood pressure, controlling for sociodemographic variables.

```{r}

```

What do the model parameters mean?
